Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-05-01 paper.tex
Updates from Overleaf

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\infdiv}{D\infdivx}
\newcommand{\future}{\overrightarrow{X}}
\newcommand{\past}{\overleftarrow{X}}
\newcommand{\key}{\textbf}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}
\newcommand{\finitefuture}{X_{0\dots M-1}}
\newcommand{\finitepast}{X_{-M\dots -1}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The predictive rate-distortion curve quantifies the tradeoff between compressing information about the past of a stochastic process and predicting its future accurately.
Existing estimation methods for this curve work by clustering finite sequences of observations or by utilizing analytically known causal states.
Neither type of approach scales to processes such as natural languages, which have large alphabets and long dependencies, and where the causal states are not known analytically.
We describe Neural Predictive Rate Distortion (NPRD), an estimation method that scales to such processes, leveraging the universal approximation capabilities of neural networks.
Taking only time series data as input, the method computes a variational bound on the predictive rate-distortion curve.
We validate the method on processes where predictive rate-distortion is analytically known.
As an application, we provide bounds on the predictive rate-distortion of natural language, improving on bounds provided by clustering sequences.
Based on the results, we argue that the predictive rate-distortion curve is more useful than the usual notion of statistical complexity for characterizing highly complex processes such as natural language.}

% Keywords
\keyword{Predictive rate-distortion; Natural language; Information bottleneck; Neural variational inference (list three to ten pertinent keywords specific to the article, yet reasonably common within the subject discipline.)}

@@

This motivates the study of models that try to extract those key features of past observations that are most relevant to predicting the future.
%This results in a tradeoff between the cost of representing past observations and predicting the future well.
%This tradeoff has been formalized and computationally studied in a range of prior studies.
A general information-theoretic framework for this problem is provided by Predictive Rate Distortion~~\citep{marzen-predictive-2016}, also known as the Past-Future Information Bottleneck~\citep{creutzig-past-future-2009}.
The Predictive Rate Distortion tradeoff seeks to find an encoding of past observations that is maximally informative about future observations, while satisfying a bound on the amount of information that has to be stored.
More formally, this framework trades off prediction loss in the future, formalized as cross-entropy, with the cost of representing the past, formalized as the mutual information between the past observations and the compressed representations of the past.
Due to its information-theoretic nature, this framework is extremely general and applies to processes across vastly different domains.
It has been applied to linear dynamical systems \citep{creutzig-past-future-2009,amir2015past}, but is equally meaningful for discrete dynamical systems~\citep{marzen-predictive-2016}.
For biological systems that make predictions about their environment, this corresponds to placing an information-theoretic constraint on the computational cost used for conditioning actions on past observations~\citep{genewein2015bounded}.

The problem of determining encodings that optimize the Predictive Rate Distortion tradeoff has been solved for certain specific kinds of dynamics, namely for linear dynamic systems~\citep{creutzig-past-future-2009} and for processes whose predictive dynamic is represented exactly by a known, finite Hidden Markov Model~\citep{marzen-predictive-2016}.
However, real-world processes are often more complicated.
When the dynamics are known, a representing Hidden Markov Model may still be extremely large or even infinite, making general-purpose automated computation difficult.
Even more importantly, the underlying dynamics are often not known exactly.
An organism typically does not have access to the exact dynamics of its surroundings.
Similarly, the exact distribution of sentences in a natural language is not known exactly, precluding the application of methods that require an exact model.
Such processes are typically only available implicitly, through a finite sample of trajectories.

Optimal Causal Estimation (OCE, \citet{still-optimal-2010}) addresses the problem of estimating Predictive Rate Distortion from a finite sample of observation trajectories.
It does so by constructing a matrix of observed frequencies of different pairs of past and future observations.
However, this method faces a series of challenges~\citep{marzen-predictive-2016}.
One is the curse of dimensionality:
Modeling long dependencies requires storing an exponential number of observations, which quickly becomes intractable for current computation methods.
This exponential growth is particularly problematic when dealing with processes with a large state space.
For instance, the number of distinct words in a human language as found in large-scale text data easily exceeds $10^5$, making storing and manipulating counts of longer word sequences very challenging.
A second challenge is that of overfitting:
When deploying a predictive model constructed via OCE on new data to predict upcoming observations, such a model can only succeed when the past sequences occurred in the sample that OCE was applied to.
This is because OCE relies on counts of full past and future observation sequences; it has no means to generalize to unseen past sequences.

@@

We propose Neural Predictive Rate Distortion (NPRD) to estimate predictive rate distortion when only a finite set of sample trajectories is given.
We use neural networks both to encode past observations into a summary code, and to predict future observations from it.
The universal function approximation capabilities of neural networks enable such networks to capture complex dynamics, with computational cost scaling only linearly with the length of observed trajectories, compared to the exponential cost of OCE.
When deploying on new data, such a neural model can generalize seamlessly to unseen sequences, and generate appropriate novel summary encodings on the fly.
Recent advances in neural variational inference~\citep{rezende-variational-2015,huang-neural-2018} allow us to construct predictive models that provide almost optimal predictive performance at a given rate, and to estimate the predictive rate distortion tradeoff from such networks.
Our method can be applied to sample trajectories in an off-the-shelf manner, without prior knowledge of the underlying process.

In Section~\ref{sec:background}, we formally introduce predictive rate distortion, and discuss related notions of predictive complexity.
In Section~\ref{sec:oce}, we describe the prior method of Optimal Causel Estimation (OCE).
In Section~\ref{sec:nprd}, we describe our method NPRD.
In Section~\ref{sec:experiments}, we validate NPRD on processes whose predictive rate distortion is analytically known, showing that it finds essentially optimal predictive models.
In Section~\ref{sec:language}, we apply NPRD to data from five languages, providing the first estimates of predictive rate-distortion of natural language.

@@

Predictive Rate-Distortion is closely related to Statistical Complexity and the $\epsilon$-machine \citep{crutchfield-inferring-1989,Grassberger1986}. 
Given a stationary process $X_t$, its \key{causal states} are the equivalence classes of semi-infinite pasts $\past$ that induce the same conditional probability over semi-infinite futures $\future$:
Two pasts $\past, \past^\prime$ belong to the same causal state if and only if $P(x_{1...M}|\past) = P(x_{1...M}|\past^\prime)$ holds for all finite sequences $x_{0...M}$ ($M \in \mathbb{N}$).
Note that this definition is not measure-theoretically rigorous; such a treatment is provided by~\citet{lohr-properties-2009}.

@@

\section{Prior Work: Optimal Causal Estimation}\label{sec:oce}

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Estimation (OCE, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts $(X_{-M \dots -1}), (\finitefuture)$:
\begin{equation}\label{eq:ib-oce}
	\max_{Z} \left( \operatorname{I}[\finitepast, Z] - \lambda \cdot \operatorname{I}[Z, \finitefuture] \right).
\end{equation}
Second, it estimates information measures directly from the observed counts of $(\finitepast), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|X_{-M \dots -1})$ that maps each observed past sequence $X_{-M...-1}$ to a distribution over a (finite) set of code words $Z$.
%From this, one obtains a corresponding decoder $P(X_{0...M}|Z)$ via Bayes' Rule.

@@

The second challenge is that OCE requires the construction of a matrix whose rows and columns are indexed by the observed past and future sequences \citep{marzen-predictive-2016}.
Depending on the topological entropy of the process, the number of such sequences can grow as $|A|^M$, where $A$ is the set of observed symbols, and processes of interest often do show this exponential growth \citep{marzen-predictive-2016}.
Drastically, in the case of natural language, $A$ contains thousands of words.
We will see that this makes applying OCE infeasible even with $M=1$.

A further challenge is that OCE is infeasible if the number of required codewords is too large, again because OCE requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $\operatorname{I}[\past, Z] > \log 10^5 \approx 11.5$ cannot be captured with OCE.

@@

We now introduce our method, Neural Predictive Rate Distortion (NPRD), to address the limitations of OCE, by using parametric function approximation:
Whereas OCE constructs a codebook mapping between observed sequences and codes, we use general-purpose function approximation estimation methods to compute the representation $Z$ from the past and to estimate a distribution over future sequences from $Z$.
In particular, we will use recurrent neural networks, which are known to provide good models of sequences from a wide range of domains; our method will be applicable to other types of function approximators also.
%The parameters of the function approximators are optimized on the observed pairs of pasts and futures.
%When deploying the resulting mapping on unseen data, the function approximator can be used to generate appropriate representations $Z$ even for sequences that did not occur in the initial data sample.

This will have two main advantages, addressing the limitations of OCE:
First, unlike OCE, function approximators can discover generalizations across similar sequences, enabling the method to calculate good codes $Z$ even for past sequences that were not seen previously.
This is of paramount importance in settings where the state space is large, such as the set of words of a natural language.
Second, the cost of storing and evaluating the function approximators will scale \emph{linearly} with the length of observed sequences both in space and in time, as opposed to the exponential memory demand of OCE.
This is crucial for modeling long dependencies.

@@

As in OCE (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: \rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read. $^M\overleftarrow{X}$}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | X_{-M\dots-1}} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | X_{-M\dots-1}}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ X_{-M\dots-1}, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

@@

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|\finitepast),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $\finitepast$ into the parameters of the distribution $P_\phi(\cdot|\finitepast)$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P(\finitefuture, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(\finitefuture|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[\finitepast, Z]$ of the code $Z \sim P_\phi(Z|\finitepast)$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

Our method will be to introduce additional functions, also expressed using function approximators, that approximate some of these intractable quantities:
First, we will use a parameterized probability distribution $q$ as an approximation to the intractable marginal $P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})$:
\begin{equation}
q(Z) \text{ approximates } P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1}).
\end{equation}
Second, to approximate the decoding distribution $P(\finitefuture|Z)$, we introduce a parameterized decoder $\psi$ that maps points $Z \in \mathbb{R}^N$  into probability distributions $P_\psi(\finitefuture|Z)$ over future observations $\finitefuture$:
\begin{equation}
P_\psi(\finitefuture|Z)  \text{ approximates }  P(\finitefuture|Z)
\end{equation}
for each code $Z \in \mathbb{R}^N$. Crucially, $P_\psi(\finitefuture|Z)$ will be easy to compute efficiently, unlike the exact decoding distribution $P(\finitefuture|Z)$.

@@

\begin{proposition} 
\label{prop:kl-bound}
The KL Divergence between $P_\phi(Z|X_{-M\dots-1})$ and $q(Z)$, averaged over pasts $X_{-M\dots-1}$, upper-bounds the rate of $Z$: % mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X_{-M\dots-1}}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots-1})}{q(Z)}\right] &=   \mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log \frac{P_\phi(Z|\finitepast)}{q(Z)}  \\
	& \geq  \mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log \frac{P_\phi(Z|\finitepast)}{P(Z)}  \\
	& = \operatorname{I}[\finitepast,Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})$.
\end{proposition}

\begin{proof}
The two equalities follow from the definition of KL Divergence and Mutual Information. To show the inequality, we again use Gibbs' inequality:
\begin{align*}
    -\mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log q(Z) = - \mathbb{E}_Z \log q(Z) \geq - \mathbb{E}_Z \log P(Z) = -\mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log P(Z).
\end{align*}
Here, equality holds if and only if $q(Z) = P(Z)$, proving the second assertion.
\end{proof}

>>>>

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\infdiv}{D\infdivx}
\newcommand{\future}{\overrightarrow{X}}
\newcommand{\past}{\overleftarrow{X}}
\newcommand{\key}{\textbf}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}
\newcommand{\finitefuture}{\overrightarrow{X}^{M}}
\newcommand{\finitepast}{^{-M}\overleftarrow{X}}%_{-M\dots -1}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The predictive rate-distortion curve quantifies the tradeoff between compressing information about the past of a stochastic process and predicting its future accurately.
Existing estimation methods for this curve work by clustering finite sequences of observations or by utilizing analytically known causal states.
Neither type of approach scales to processes such as natural languages, which have large alphabets and long dependencies, and where the causal states are not known analytically.
We describe Neural Predictive Rate-Distortion (NPRD), an estimation method that scales to such processes, leveraging the universal approximation capabilities of neural networks.
Taking only time series data as input, the method computes a variational bound on the predictive rate-distortion curve.
We validate the method on processes where predictive rate-distortion is analytically known.
As an application, we provide bounds on the predictive rate-distortion of natural language, improving on bounds provided by clustering sequences.
Based on the results, we argue that the predictive rate-distortion curve is more useful than the usual notion of statistical complexity for characterizing highly complex processes such as natural language.}

% Keywords
\keyword{Predictive rate-distortion; Natural language; Information bottleneck; Neural variational inference} % (list three to ten pertinent keywords specific to the article, yet reasonably common within the subject discipline.)

@@

This motivates the study of models that try to extract those key features of past observations that are most relevant to predicting the future.
%This results in a tradeoff between the cost of representing past observations and predicting the future well.
%This tradeoff has been formalized and computationally studied in a range of prior studies.
A general information-theoretic framework for this problem is provided by Predictive Rate-Distortion~~\citep{marzen-predictive-2016}, also known as the Past-Future Information Bottleneck~\citep{creutzig-past-future-2009}.
The Predictive Rate-Distortion tradeoff seeks to find an encoding of past observations that is maximally informative about future observations, while satisfying a bound on the amount of information that has to be stored.
More formally, this framework trades off prediction loss in the future, formalized as cross-entropy, with the cost of representing the past, formalized as the mutual information between the past observations and the compressed representations of the past.
Due to its information-theoretic nature, this framework is extremely general and applies to processes across vastly different domains.
It has been applied to linear dynamical systems \citep{creutzig-past-future-2009,amir2015past}, but is equally meaningful for discrete dynamical systems~\citep{marzen-predictive-2016}.
For biological systems that make predictions about their environment, this corresponds to placing an information-theoretic constraint on the computational cost used for conditioning actions on past observations~\citep{genewein2015bounded}.

The problem of determining encodings that optimize the Predictive Rate-Distortion tradeoff has been solved for certain specific kinds of dynamics, namely for linear dynamic systems~\citep{creutzig-past-future-2009} and for processes whose predictive dynamic is represented exactly by a known, finite Hidden Markov Model~\citep{marzen-predictive-2016}.
However, real-world processes are often more complicated.
When the dynamics are known, a representing Hidden Markov Model may still be extremely large or even infinite, making general-purpose automated computation difficult.
Even more importantly, the underlying dynamics are often not known exactly.
An organism typically does not have access to the exact dynamics of its surroundings.
Similarly, the exact distribution of sentences in a natural language is not known exactly, precluding the application of methods that require an exact model.
Such processes are typically only available implicitly, through a finite sample of trajectories.

Optimal Causal Filtering (OCF, \citet{still-optimal-2010}) addresses the problem of estimating Predictive Rate-Distortion from a finite sample of observation trajectories.
It does so by constructing a matrix of observed frequencies of different pairs of past and future observations.
However, this method faces a series of challenges~\citep{marzen-predictive-2016}.
One is the curse of dimensionality:
Modeling long dependencies requires storing an exponential number of observations, which quickly becomes intractable for current computation methods.
This exponential growth is particularly problematic when dealing with processes with a large state space.
For instance, the number of distinct words in a human language as found in large-scale text data easily exceeds $10^5$, making storing and manipulating counts of longer word sequences very challenging.
A second challenge is that of overfitting:
When deploying a predictive model constructed via OCF on new data to predict upcoming observations, such a model can only succeed when the past sequences occurred in the sample that OCF was applied to.
This is because OCF relies on counts of full past and future observation sequences; it has no means to generalize to unseen past sequences.

@@

We propose Neural Predictive Rate-Distortion (NPRD) to estimate predictive rate-distortion when only a finite set of sample trajectories is given.
We use neural networks both to encode past observations into a summary code, and to predict future observations from it.
The universal function approximation capabilities of neural networks enable such networks to capture complex dynamics, with computational cost scaling only linearly with the length of observed trajectories, compared to the exponential cost of OCF.
When deploying on new data, such a neural model can generalize seamlessly to unseen sequences, and generate appropriate novel summary encodings on the fly.
Recent advances in neural variational inference~\citep{rezende-variational-2015,huang-neural-2018} allow us to construct predictive models that provide almost optimal predictive performance at a given rate, and to estimate the predictive rate-distortion tradeoff from such networks.
Our method can be applied to sample trajectories in an off-the-shelf manner, without prior knowledge of the underlying process.

In Section~\ref{sec:background}, we formally introduce predictive rate-distortion, and discuss related notions of predictive complexity.
In Section~\ref{sec:oce}, we describe the prior method of Optimal Causel Filtering (OCF).
In Section~\ref{sec:nprd}, we describe our method NPRD.
In Section~\ref{sec:experiments}, we validate NPRD on processes whose predictive rate-distortion is analytically known, showing that it finds essentially optimal predictive models.
In Section~\ref{sec:language}, we apply NPRD to data from five languages, providing the first estimates of predictive rate-distortion of natural language.

@@

Predictive Rate-Distortion is closely related to Statistical Complexity and the $\epsilon$-machine \citep{crutchfield-inferring-1989,Grassberger1986}. 
Given a stationary process $X_t$, its \key{causal states} are the equivalence classes of semi-infinite pasts $\past$ that induce the same conditional probability over semi-infinite futures $\future$:
Two pasts $\past, \past^\prime$ belong to the same causal state if and only if $P(x_{1...k}|\past) = P(x_{1...k}|\past^\prime)$ holds for all finite sequences $x_{0...k}$ ($k \in \mathbb{N}$).
Note that this definition is not measure-theoretically rigorous; such a treatment is provided by~\citet{lohr-properties-2009}.

@@

\section{Prior Work: Optimal Causal Filtering}\label{sec:OCF}

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Filtering (OCF, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts, i.e., pairs of finite past contexts $(\finitepast := X_{-M}\dots X_{-1})$ and future contexts $(\finitefuture := X_{0}\dots X_{M-1})$ of some finite length $M$.\footnote{It is not crucial that past and future contexts have the same lengths, and indeed \citet{still-optimal-2010} do not assume this. We do assume equal length throughout this paper for consistency with our experiments, though nothing depends on this.}
The PRD objective (\ref{eq:ib}) then becomes (\ref{eq:ib-oce}), aiming to predict length-$M$ finite futures from summary codes $Z$ of length-$M$ finite pasts:
\begin{equation}\label{eq:ib-oce}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
Second, OCF estimates information measures directly from the observed counts of $(\finitepast), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib-oce}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|\finitepast)$ that maps each observed past sequence $\finitepast$ to a distribution over a (finite) set of code words $Z$.

@@

The second challenge is that OCF requires the construction of a matrix whose rows and columns are indexed by the observed past and future sequences \citep{marzen-predictive-2016}.
Depending on the topological entropy of the process, the number of such sequences can grow as $|A|^M$, where $A$ is the set of observed symbols, and processes of interest often do show this exponential growth \citep{marzen-predictive-2016}.
Drastically, in the case of natural language, $A$ contains thousands of words.
We will see that this makes applying OCF infeasible even with $M=1$.

A further challenge is that OCF is infeasible if the number of required codewords is too large, again because OCF requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $\operatorname{I}[\past, Z] > \log 10^5 \approx 11.5$ cannot be captured with OCF.

@@

We now introduce our method, Neural Predictive Rate-Distortion (NPRD), to address the limitations of OCF, by using parametric function approximation:
Whereas OCF constructs a codebook mapping between observed sequences and codes, we use general-purpose function approximation estimation methods to compute the representation $Z$ from the past and to estimate a distribution over future sequences from $Z$.
In particular, we will use recurrent neural networks, which are known to provide good models of sequences from a wide range of domains; our method will be applicable to other types of function approximators also.
%The parameters of the function approximators are optimized on the observed pairs of pasts and futures.
%When deploying the resulting mapping on unseen data, the function approximator can be used to generate appropriate representations $Z$ even for sequences that did not occur in the initial data sample.

This will have two main advantages, addressing the limitations of OCF:
First, unlike OCF, function approximators can discover generalizations across similar sequences, enabling the method to calculate good codes $Z$ even for past sequences that were not seen previously.
This is of paramount importance in settings where the state space is large, such as the set of words of a natural language.
Second, the cost of storing and evaluating the function approximators will scale \emph{linearly} with the length of observed sequences both in space and in time, as opposed to the exponential memory demand of OCF.
This is crucial for modeling long dependencies.

@@

As in OCF (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: %\rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read. $^M\overleftarrow{X}$}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ \finitepast, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

@@

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|\finitepast),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $\finitepast$ into the parameters of the distribution $P_\phi(\cdot|\finitepast)$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} \\
	&= \frac{\E_{\finitepast} P(\finitefuture, Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} \\
	&=^{(*)} \frac{\E_{\finitepast}   P(\finitefuture|\finitepast)     P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition $Z \bot \finitefuture | \finitepast$.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[\finitepast, Z]$ of the code $Z \sim P_\phi(Z|\finitepast)$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

Our method will be to introduce additional functions, also expressed using function approximators, that approximate some of these intractable quantities:
First, we will use a parameterized probability distribution $q$ as an approximation to the intractable marginal $P(Z) = \E_{\finitepast} P_\phi(Z|\finitepast)$:
\begin{equation}
q(Z) \text{ approximates } P(Z) = \E_{\finitepast} P_\phi(Z|\finitepast).
\end{equation}
Second, to approximate the decoding distribution $P(\finitefuture|Z)$, we introduce a parameterized decoder $\psi$ that maps points $Z \in \mathbb{R}^N$  into probability distributions $P_\psi(\finitefuture|Z)$ over future observations $\finitefuture$:
\begin{equation}
P_\psi(\finitefuture|Z)  \text{ approximates }  P(\finitefuture|Z)
\end{equation}
for each code $Z \in \mathbb{R}^N$. Crucially, $P_\psi(\finitefuture|Z)$ will be easy to compute efficiently, unlike the exact decoding distribution $P(\finitefuture|Z)$.

@@

\begin{proposition} 
\label{prop:kl-bound}
The KL Divergence between $P_\phi(Z|\finitepast)$ and $q(Z)$, averaged over pasts $\finitepast$, upper-bounds the rate of $Z$: % mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] &=   \mathbb{E}_{\finitepast} \mathbb{E}_{Z|\finitepast} \log \frac{P_\phi(Z|\finitepast)}{q(Z)}  \\
	& \geq  \mathbb{E}_{\finitepast} \mathbb{E}_{Z|\finitepast} \log \frac{P_\phi(Z|\finitepast)}{P(Z)}  \\
	& = \operatorname{I}[\finitepast,Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{\finitepast} P_\phi(Z|\finitepast)$.
\end{proposition}

\begin{proof}
The two equalities follow from the definition of KL Divergence and Mutual Information. To show the inequality, we again use Gibbs' inequality:
\begin{align*}
    -\mathbb{E}_{\finitepast} \mathbb{E}_{Z|\finitepast} \log q(Z) = - \mathbb{E}_Z \log q(Z) \geq - \mathbb{E}_Z \log P(Z) = -\mathbb{E}_{\finitepast} \mathbb{E}_{Z|\finitepast} \log P(Z).
\end{align*}
Here, equality holds if and only if $q(Z) = P(Z)$, proving the second assertion.
\end{proof}

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:prd-loss}) in a way amenable to using function approximators.
Recall the predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\tag{\ref{eq:prd-loss}}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \operatorname{I}[ \finitepast, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
By the two Propositions, this objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-loss}) by setting $Z$ to be the code generated by the encoder $\phi$.
The second proposition guarantees that, for a solution to  (\ref{eq:prd-loss}), the rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, \finitepast] =   \mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[\finitefuture|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. %H[\finitefuture|\finitepast] % \operatorname{H}[\finitefuture]
\end{equation}
Note that the right-hand-sides of (\ref{eq:eq1-new}-\ref{eq:eq2-new}) can both be estimated efficiently using Monte Carlo samples from $Z \sim P_\phi(\finitepast)$.

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction loss for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, \finitepast] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[\finitefuture|\finitepast]
  \operatorname{H}[\finitefuture|Z] \leq   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. % \operatorname{H}[\finitefuture]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and---if the family of approximators is sufficiently rich---can come close to turning these into the equalities (\ref{eq:eq1-new}-\ref{eq:eq2-new}).
